Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
refactoring noah code

\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\usepackage{graphicx}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

In~Viterbi's equations,
the probability in each state is a simple function of the probability
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm
which starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
We~implemented this algorithm, storing the probabilities in an array.
\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
and what??) and why do we care?}
The cost is
$O(|N|\times|\alignwidth|)$;
in our work, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

\emph{[WHAT WAS UNSATISFACTORY ABOUT THE FIRST IMPLEMENTATION]}

We next wrote a top-down dynamic-programming 
version, which began with the last node of the model and the last observation, 
and applied structural recursion down to the base cases involving the first 
node or first observation (or both).
The general case of this top-down, structurally recursive implementation closely
resembles the Viterbi equations above.
The match equation is thus:
\begin{smallverbatim}
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

>>>>

\newcommand\naive{na\"\i ve}
\usepackage{amsmath}
\usepackage{array}
\usepackage{listings}
\usepackage{graphicx}
\lstset{
    tabsize = 2,
    basicstyle = \ttfamily,
    language = Haskell
    }

@@

In~Viterbi's equations,
the probability in each state is a simple function of the probability
in its predecessor states, 
and all probabilities can be computed by a classic dynamic-programming
algorithm
which starts at the begin state,
computes probabilites in nodes $1$~through~$\alignwidth$ in
succession, and terminates at the end state.
One of us implemented this algorithm, storing the probabilities in an array.
\remark{Why is the number of dimensions 3 and not 2 (indices $i$, $j$,
and what??) and why do we care?}
The cost was
$O(|N|\times|\alignwidth|)$;
in our work, $\alignwidth$~and~$N$ range from several hundred to a few
thousand.

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
The negated log probability is called a \texttt{Score}.
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{smallverbatim}
viterbi :: State -> Int -> Int -> (Score, StatePath)
viterbi Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
ALTERNATE VERSION FOLLOWS
\begin{smallverbatim}
viterbi :: WHAT IS MY TYPE????
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
